Contextuality and Nonlocality in Discrete Quantum Theory

### 2015-12-22 DQC3.tex
Typed the proof when d = 3 for 4.4 A No-Go Theorem

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{amsfonts}
\usepackage{color}
\usepackage{stmaryrd}\usepackage{dsfont}
\usepackage[usenames,dvipsnames]{xcolor}

@@

\usepackage{babel}

@@

\begin{theorem}
\label{theorem1}
There is no map $\dpt\colon \ffzd{p^2}\times \ffzd{p^2}
\longrightarrow [0,1]$ satisfying properties ({\bf i}), ({\bf ii}),
and ({\bf iii}) when $d> 3$.
\end{theorem}
%
\begin{proof}
  Assume such a map $\dpt$ satisfying all three properties exists.
  Consider the state
\[
\ket{\Psi}=\alpha\big(\ket{0} + \ket{1} + \ket{2} + \beta\ket{3}\big)
\]
expressed in terms of orthonormal vectors, where $\beta\in \ff{p^2}$
is chosen so that $\left|\beta\right|^2=p-1$ and $\alpha\in\ff{p^2}$
is a constant such that $\ket{\Psi}$ is normalized.  Let
$\ket{\Phi}= \ket{1} + \ket{2} + \beta\ket{3}$. Note that $\ket{\Phi}$
is normalized in $\ff{p}$.  We can write our original state
$\ket{\Psi}$ as $\ket{\Psi}=\alpha\big(\ket{0}+\ket{\Phi}\big)$ The
vectors $\ket{0}$ and $\ket{\Phi}$ are orthonormal and so are part of
some larger orthonormal basis. Thus by properties ({\bf i}) and ({\bf
  ii}) we have
\begin{equation}
1= \dpt\big(\ket{\Psi},\ket{0}\big) +  \dpt\big(\ket{\Psi},\ket{\Phi}\big). \label{prob1}
\end{equation}
Let $U$ be the unitary map that simply permutes the orthonormal
vectors $\ket{0}$ and $\ket{\Phi}$.  Note that $\ket{\Psi}$ is
invariant under $U$, so invoking property ({\bf iii}), we have
\[
\dpt\big(\ket{\Psi},\ket{0}\big)=\dpt\big(U\ket{\Psi},U\ket{0}\big) = 
   \dpt\big(\ket{\Psi},\ket{\Phi}\big) \ .
\]
This along with equation (\ref{prob1}) gives
\[
1= \dpt\big(\ket{\Psi},\ket{0}\big) +  \dpt\big(\ket{\Psi},\ket{0}\big),
\]
so
\begin{equation}
\label{prob2}
\dpt\big(\ket{\Psi},\ket{0}\big)=\dpt\big(\ket{\Psi},
   \ket{\Phi}\big)=\frac{1}{2} \in [0,1] \ .
\end{equation}

For $i=0,1,2$ let $U_i$ be the unitary map that permutes the basis
vectors $\ket{0}$ and $\ket{i}$ and acts as the identity for the rest.
Note that $\ket{\Psi}$ is invariant under $U_i$. So again by property
({\bf iii}) we have
\[
\dpt\big(\ket{\Psi},\ket{i}\big) =\dpt\big(U_i\ket{\Psi},U_i\ket{0}\big)=
\dpt\big(\ket{\Psi},\ket{0}\big) = \frac{1}{2}
\]
for $i=0,1,2$.  These probability assignments are inconsistent with
the requirement that the probabilities for orthogonal outcomes add up
to 1. By property ({\bf i})
\begin{eqnarray*}
1 & = & \dpt\big(\ket{\Psi},\ket{0}\big)+\dpt\big(\ket{\Psi},\ket{1}\big)+\dpt\big(\ket{\Psi},\ket{2}\big)+\dpt\big(\ket{\Psi},\ket{3}\big)\\
 & = & \frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\dpt\big(\ket{\Psi},\ket{3}\big)\\
 & \geq & \frac{3}{2},
\end{eqnarray*}
so we have a contradiction. The assumption of a satisfactory map
$\dpt$ is thus false. $\Box$
\end{proof}
(It should be pointed out that the argument above explicitly uses
states from a vector space of four dimensions or larger, and thus does
not hold if the largest vector space one can consider has dimension
$d\le 3$.  But it does hold for any system that, for example, contains
more than a single qubit.)

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{amsfonts}
\usepackage{color}
\usepackage{stmaryrd}\usepackage{dsfont}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{enumitem}		% customizable list environments
\newlength{\lyxlabelwidth}      % auxiliary length 
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
\ifx\proof\undefined
\newenvironment{proof}[1][\protect\proofname]{\par
\normalfont\topsep6\p@\@plus6\p@\relax
\trivlist
\itemindent\parindent
\item[\hskip\labelsep\scshape #1]\ignorespaces
}{%
\endtrivlist\@endpefalse
}
\providecommand{\proofname}{Proof}
\fi

@@

\usepackage{babel}
\providecommand{\theoremname}{Theorem}

@@

\begin{thm}
\label{theorem1} There is no map $\dpt\colon\ffzd{p^{2}}\times\ffzd{p^{2}}\longrightarrow[0,1]$
satisfying properties ({\bf i}), ({\bf ii}), and ({\bf iii})
when $d>3$. \end{thm}
%
\begin{proof}
Assume such a map $\dpt$ satisfying all three properties exists.
Consider the state 
\[
\ket{\Psi}=\alpha\big(\ket{0}+\ket{1}+\ket{2}+\beta\ket{3}\big)
\]
expressed in terms of orthonormal vectors, where $\beta\in\ff{p^{2}}$
is chosen so that $\left|\beta\right|^{2}=p-1$ and $\alpha\in\ff{p^{2}}$
is a constant such that $\ket{\Psi}$ is normalized. Let $\ket{\Phi}=\ket{1}+\ket{2}+\beta\ket{3}$.
Note that $\ket{\Phi}$ is normalized in $\ff{p}$. We can write our
original state $\ket{\Psi}$ as $\ket{\Psi}=\alpha\big(\ket{0}+\ket{\Phi}\big)$
The vectors $\ket{0}$ and $\ket{\Phi}$ are orthonormal and so are
part of some larger orthonormal basis. Thus by properties ({\bf
i}) and ({\bf ii}) we have 
\begin{equation}
1=\dpt\big(\ket{\Psi},\ket{0}\big)+\dpt\big(\ket{\Psi},\ket{\Phi}\big).\label{prob1}
\end{equation}
Let $U$ be the unitary map that simply permutes the orthonormal vectors
$\ket{0}$ and $\ket{\Phi}$. Note that $\ket{\Psi}$ is invariant
under $U$, so invoking property ({\bf iii}), we have 
\[
\dpt\big(\ket{\Psi},\ket{0}\big)=\dpt\big(U\ket{\Psi},U\ket{0}\big)=\dpt\big(\ket{\Psi},\ket{\Phi}\big)\ .
\]
This along with equation (\ref{prob1}) gives 
\[
1=\dpt\big(\ket{\Psi},\ket{0}\big)+\dpt\big(\ket{\Psi},\ket{0}\big),
\]
so 
\begin{equation}
\dpt\big(\ket{\Psi},\ket{0}\big)=\dpt\big(\ket{\Psi},\ket{\Phi}\big)=\frac{1}{2}\in[0,1]\ .\label{prob2}
\end{equation}
\par For $i=0,1,2$ let $U_{i}$ be the unitary map that permutes
the basis vectors $\ket{0}$ and $\ket{i}$ and acts as the identity
for the rest. Note that $\ket{\Psi}$ is invariant under $U_{i}$.
So again by property ({\bf iii}) we have 
\[
\dpt\big(\ket{\Psi},\ket{i}\big)=\dpt\big(U_{i}\ket{\Psi},U_{i}\ket{0}\big)=\dpt\big(\ket{\Psi},\ket{0}\big)=\frac{1}{2}
\]
for $i=0,1,2$. These probability assignments are inconsistent with
the requirement that the probabilities for orthogonal outcomes add
up to 1. By property ({\bf i}) 
\begin{eqnarray*}
1 & = & \dpt\big(\ket{\Psi},\ket{0}\big)+\dpt\big(\ket{\Psi},\ket{1}\big)+\dpt\big(\ket{\Psi},\ket{2}\big)+\dpt\big(\ket{\Psi},\ket{3}\big)\\
 & = & \frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\dpt\big(\ket{\Psi},\ket{3}\big)\\
 & \geq & \frac{3}{2},
\end{eqnarray*}
so we have a contradiction. The assumption of a satisfactory map $\dpt$
is thus false. 
\end{proof}
(It should be pointed out that the argument above explicitly uses
states from a vector space of four dimensions or larger, and thus does
not hold if the largest vector space one can consider has dimension
$d\le 3$.  But it does hold for any system that, for example, contains
more than a single qubit.)

\yutsung{I am going to prove a no-go theorem for $d=3$ here with
an extra property
\begin{description}
\item [{(iv)}] $\dpt(\ket{\Phi},\ket{\Psi})=\dpt(\lambda\ket{\Phi},\ket{\Psi})=\dpt(\ket{\Phi},\lambda\ket{\Psi})$
where $\lambda\in\ff{p^{2}}^{*}$. 
\end{description}
This property means that the phase difference and normalization will
not effect the probability assignment. Although this property will
not be used explicitly, considering non-normalized states may simplify
the proof. 
\begin{thm}
\label{theorem1-1} There is no map $\dpt\colon\ffzd{p^{2}}\times\ffzd{p^{2}}\longrightarrow[0,1]$
satisfying properties ({\bf i}), ({\bf ii}), ({\bf iii}),
and ({\bf iv}) when $d\ge3$ and $p>3$.\end{thm}
}

\yutsung{
\begin{proof}
Assume such a map $\dpt$ satisfying all four properties exists. Consider
the orthonormal basis and the non-normalized state 
\begin{eqnarray*}
\ket{0} & = & \left(\begin{array}{ccc}
1 & 0 & 0\end{array}\right)^{T}\\
\ket{1} & = & \left(\begin{array}{ccc}
0 & 1 & 0\end{array}\right)^{T}\\
\ket{2} & = & \left(\begin{array}{ccc}
0 & 0 & 1\end{array}\right)^{T}\\
\ket{+} & = & \ket{0}+\ket{1}+\ket{2}\textrm{ ,}
\end{eqnarray*}
where $\ket{+}$ is not zero-normed when $p>3$. For $i=1,2$, let
$\mathbf{U}_{i}$ be the unitary map that permutes the basis vectors
$\ket{0}$ and $\ket{i}$ and acts as the identity for the rest. Note
that $\ket{+}$ is invariant under $\mathbf{U}_{i}$. So by property
({\bf iii}) we have 
\[
\dpt\big(\ket{+},\ket{i}\big)=\dpt\big(\mathbf{U}_{i}\ket{+},\mathbf{U}_{i}\ket{0}\big)=\dpt\big(\ket{+},\ket{0}\big)
\]
for $i=1,2$. By property ({\bf i}) 
\begin{eqnarray*}
\dpt\big(\ket{+},\ket{0}\big)+\dpt\big(\ket{+},\ket{1}\big)+\dpt\big(\ket{+},\ket{2}\big) & = & 1
\end{eqnarray*}
so 
\begin{equation}
\dpt\big(\ket{+},\ket{i}\big)=\frac{1}{3}\in[0,1]\textrm{ .}\label{prob2-1-1}
\end{equation}

Let 
\begin{eqnarray*}
\mathbf{U}' & = & \left(\begin{array}{ccc}
\alpha & \alpha & 0\\
-\alpha & \alpha & 0\\
0 & 0 & 1
\end{array}\right)\\
\ket{\Phi} & = & 2\alpha\ket{0}+\ket{2}=\mathbf{U}'\ket{+}
\end{eqnarray*}
where $\alpha\in\ff{p^{2}}$ is chosen so that $2\alpha^{*}\alpha=1$.
Thus by property ({\bf iii}), we have
\begin{eqnarray*}
\dpt\big(\ket{\Phi},\ket{2}\big) & = & \dpt\big(\mathbf{U}'\ket{+},\mathbf{U}'\ket{2}\big)=\dpt\big(\ket{+},\ket{2}\big)\textrm{ .}
\end{eqnarray*}
Also, properties ({\bf i}) and ({\bf ii}) imply 
\begin{eqnarray*}
1 & = & \dpt\big(\ket{\Phi},\ket{0}\big)+\dpt\big(\ket{\Phi},\ket{2}\big)\textrm{ .}
\end{eqnarray*}
This along with equation (\ref{prob2-1-1}) gives 
\begin{equation}
\dpt\big(\ket{\Phi},\ket{0}\big)=\frac{2}{3}\in[0,1]\textrm{ .}\label{eq:prob23}
\end{equation}

\end{proof}
}

\yutsung{
\begin{proof}
Finally, consider 
\begin{eqnarray*}
\mathbf{U}'' & = & \left(\begin{array}{ccc}
1 & 0 & 0\\
0 & \beta^{*} & 2\alpha\\
0 & 2\alpha^{*} & \beta
\end{array}\right)\\
\ket{\Psi} & = & 2\alpha\left(\ket{0}+\ket{1}\right)+\beta\ket{2}=\mathbf{U}''\ket{\Phi}\textrm{ ,}
\end{eqnarray*}
where $\beta\in\ff{p^{2}}$ is chosen so that $\beta^{*}\beta=-1$.
Because $\ket{\Psi}$ is invariant under $\mathbf{U}_{1}$, invoking
property ({\bf iii}) with equation (\ref{eq:prob23}), we have
\begin{eqnarray*}
 &  & \dpt\big(\ket{\Psi},\ket{1}\big)=\dpt\big(\mathbf{U}_{1}\ket{\Psi},\mathbf{U}_{1}\ket{1}\big)\\
 & = & \dpt\big(\ket{\Psi},\ket{0}\big)=\dpt\big(\mathbf{U}''\ket{\Phi},\mathbf{U}''\ket{0}\big)\\
 & = & \dpt\big(\ket{\Phi},\ket{0}\big)=\frac{2}{3}
\end{eqnarray*}
These probability assignments are inconsistent with the requirement
that the probabilities for orthogonal outcomes add up to 1. By property
({\bf i}) 
\begin{eqnarray*}
1 & = & \dpt\big(\ket{\Psi},\ket{0}\big)+\dpt\big(\ket{\Psi},\ket{1}\big)+\dpt\big(\ket{\Psi},\ket{2}\big)\\
 & = & \frac{2}{3}+\frac{2}{3}+\dpt\big(\ket{\Psi},\ket{2}\big)\\
 & \geq & \frac{4}{3}\textrm{ ,}
\end{eqnarray*}
so we have a contradiction. The assumption of a satisfactory map $\dpt$
is thus false. 
\end{proof}
}
